\LARGE \bf Planning with Attitude

### 2020-10-10 paper/main.tex
Update barrell roll example and refine paper

% Math
\usepackage{physics}
\usepackage{siunitx}
\sisetup{output-exponent-marker=\ensuremath{\mathrm{e}}}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}

@@

\begin{abstract}
Planning trajectories for floating-base robotic systems that experience
large attitude changes is challenging due to the nontrivial group structure of 3D
rotations. This paper introduces a powerful and accessible approach for 
optimization-based planning on the space of rotations based on vector calculus and
linear algebra. We demonstrate the effectiveness of the approach by adapting Newton's
method to solve the canonical Wahba's problem, and modifying the ALTRO solver for 
constrained nonlinear trajectory optimization to plan directly on the space of unit 
quaternions, achieving superior convergence on problems involving significant changes 
in attitude.
\keywords{
    motion planning and control, quaternions, optimal control, linear quadratic regulator
}
\end{abstract}

@@

    \begin{itemize}
        \item A unified approach to quaternion differential calculus entirely based on
        standard vector calculus and linear algebraic operations control of systems with
        quaternion states
        \item A fast and efficient solver for trajectory optimization problems with 
        attitude dynamics and nonlinear constraints that correctly accounts for the group
        structure of 3D rotations during the solve
    \end{itemize}

@@

    Given known vectors in some inertial frame, ${}^N v_i$, and measurements of these
    vectors in some body fixed frame, ${}^B v_i$, our goal is to determine the relative
    rotation from the body to inertial frames ${}^N q^N$, expressed as a quaternion. We
    can define Wahba's loss function as the following: 
    \begin{align} \label{eq:wahba_loss}
        L &= \sum_i w_i \norm{{}^N v_i - {}^N A(q)^B \,\, ^Bv_i }_2^2. = \norm{r_i(q)}_2^2 
    \end{align} 
    where $r_i(q)$ is the residual vector.

    We can solve for ${}^N q^N$ using a nonlinear least squares method minimizing
    Wahba's loss function:
    \begin{mini*}
        {q}{ \norm{r(q)}_2^2 }{}{}
        \addConstraint { q\in SO(3).}
    \end{mini*}

@@

    Alternatively, if we instead minimize with respect to a differential quaternion $\phi$,
    we adapt the algorithm by simply ``correcting'' our Jacobian $J$ using 
    \eqref{eq:quat_jacobian}:
    \begin{equation} 
        \bar{J} = \pdv{r(q \otimes \varphi{\phi})}{\phi} = \pdv{r(q)}{q} G(q).
    \end{equation}
    We obtain our step---this time in the actual tangent space---as before:
    $\phi = (\bar{J}^T \bar{J})^{-1} \bar{J}^T)$. To obtain our next iterate, we ``add'' the
    step using the correct notion of composition for the group: 
    $q_{k+1} = L(q_k) \varphi(\phi)$. This ``multiplicative'' Newton algorithm is 
    summarized in Algorithm \ref{alg:mgn}.

    \begin{algorithm} 
    	\begin{algorithmic}[1]
    		\caption{Multiplicative Gauss-Newton Method}\label{alg:mgn}
    		\State $k = 0$
    		\While{significant progress}
    		    \State $\bar{J} = \pdv{r(q_k)}{q} G(q_k)$ \Comment{quaternion adjusted Jacobian}
    		    \State $ \phi = -(\bar{J}^T \bar{J})^{-1} \bar{J}^T r(q_k)$ 
    		    \State $q_{k+1} = L(q_k) \varphi(\phi)$ \Comment{apply step multiplicatively}
    		    \State $k = k + 1$
    		\EndWhile
    	\end{algorithmic}
    \end{algorithm}

@@

\section{Trajectory Optimization on $\R^n \times SO(3)$} \label{sec:trajopt}
    Here we outline the modifications to the ALTRO solver \cite{howell2019altro}, to
    solve trajectory optimization problems for rigid bodies, which extends easily to
    arbitrary systems in $R^n \times SO(3)$. ALTRO is an efficient solver for constrained
    nonlinear optimization problems that uses iterative LQR (iLQR) with an augmented
    Lagrangian framework.

@@

    With this expansion, we calculate the expansion of the ``action-value function'' $Q(x,u)$
    as normal:
    \begin{align}
        Q_{xx} &= \ell_{xx} + A_{k}^T P_{k+1} A_{k} \label{Qxx_exp}\\
        Q_{uu} &= \ell_{uu} + B_{k}^T P_{k+1} B_{k} \label{Quu_exp}\\
        Q_{ux} &= \ell_{ux} + B_{k}^T P_{k+1} A_{k} \label{Qux_exp}\\
        Q_x &= \ell_x  + A_{k}^T p_{k+1} \label{Qx_exp}\\
        Q_u &= \ell_u + B_{k}^T p_{k+1} \label{Qu_exp},
    \end{align}
    from which we can calculate the quadratic expansion of the cost-to-go 
    $P_k \in \R^{12 \times 12}$, $p_k \in \R^12$, and optimal linearized feedback gains 
    $K_k \in \R^{m \times 12}$ and $d_k \in \R^m$ by starting at the terminal state and
    resursing backward in time along the trajectory during the ``backward pass'' of the iLQR
    algorithm. During the ``forward pass'', the dynamics are simulated forward in time using
    the feedback gains computed during the backward pass. At each time step, the control is 
    calculated using the linear feedback controller: 
    \begin{equation} \label{eq:mlqr_control}
        u_k = K_k \delta x_k + \bar{u}_k.
    \end{equation}
    where $\bar{u}_k$ is the control value from the previous iteration, and $\delta x$ is 
    computed using \eqref{eq:state_error}, with $x_k$ the current state estimate and $\bar{x}_k$
    the state from the previous iteration. The rest of the algorithm is left unchanged. 
    For more details on the ALTRO algorithm, the reader is encouraged to refer to the original
    paper \cite{howell2019altro}.

@@

\section{Experiments} \label{sec:experiments}
    In this section we present several trajectory optimization problems for systems that
    undergo large changes in attitude: an airplane barrell roll, a quadrotor flip, and a
    satellite with flexible modes that must slew to a new orientation while avoiding a 
    ``keep out'' orientation zone. All results were run on a desktop computer with an 
    AMD Ryzen 2950x processor with 40 GB of RAM. 
    All problems are run using ALTRO, first without any of the modification presented 
    in the current paper, labeled ``original'', and then using the modifications listed
    in Sec. \ref{sec:trajopt} and the geodesic cost funtion described in Sec. \ref{sec:geodesic}.
    Timing results are summarized in Table \ref{tab:timing_results}.
    All experiments were solved to a constraint satisfaction tolerance of $10^{-5}$.
    Code for all experiments is available on 
    \href{https://github.com/RoboticExplorationLab/PlanningWithAttitude}
    {GitHub\footnote{\url{https://github.com/RoboticExplorationLab/PlanningWithAttitude}}}.

@@

        We also compared using a quadratic cost on the 3-dimensional quaternion error
        described in Section \ref{sec:error_quadratic} to the geodesic cost function in
        Section \ref{sec:geodesic}. The geodesic cost was more efficient, both in term of 
        iterations and computation time, taking only 41 iterations and

@@

	    Figure \ref{fig:quad_flip} shows snapshots of the trajectory as generated using
        ALTRO. The original version of ALTRO, even after significant tuning efforts,
        would not converge to the desired solution. It instead got halfway and then ``unwound''
        360 degrees and then continued rotating to the final orientation. This behavior
        is common and expected when attempting optimization that does not properly account 
        for the grou structure of rotations.

            \begin{figure}[t]
                \centering
                \includegraphics[height=5.2cm,trim={0 0cm 0 0},clip]{figures/quadflip.png}
                \caption{Snapshots of the quadrotor flip trajectory. The
                    red-colored quadrotors represent the state near t=0 s and the
                    green-colored quadrotors represent the state near t=5.0 s
                }
                \label{fig:quad_flip}
            \end{figure}

@@

    With the modifications presented, ALTRO can solve problems few other methods for 
    trajectory optimization can. Many state-of-the-art methods such as direct 
    collocation or sequential convex programming rely on commercial, proprietary, or 
    general-purpose solvers whose internal numerics often abstracted away from the user. 
    By exploiting both the unique Markovian structure of the trajectory optimization problem
    and the group structure of rotations, ALTRO will likely be able to solve new classes of 
    problems with at or near real-time performance. Future work will focus on continued 
    refinement of the implementation and benchmarking against current state-of-the-art 
    methods for trajectory optimization.

>>>>

% Math
\usepackage{physics}
\usepackage{siunitx}
\sisetup{output-exponent-marker=\ensuremath{\mathrm{e}}}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}

@@

\begin{abstract}
Planning trajectories for floating-base robotic systems that experience
large attitude changes is challenging due to the nontrivial group structure of 3D
rotations. This paper introduces a powerful and accessible approach for 
optimization-based planning on the space of rotations based on vector calculus and
linear algebra. We demonstrate the effectiveness of the approach by adapting Newton's
method to solve the canonical Wahba's problem, and modifying the ALTRO solver for 
constrained nonlinear trajectory optimization to plan directly on the space of unit 
quaternions, achieving superior convergence on problems involving significant changes 
in attitude.
\keywords{motion planning and control, quaternions, optimal control, 
linear quadratic regulator
}
\end{abstract}

@@

    \begin{itemize}
        \item A unified approach to quaternion differential calculus entirely based on
        standard vector calculus and linear algebraic operations, and
        \item a fast and efficient solver for trajectory optimization problems with 
        attitude dynamics and nonlinear constraints that correctly accounts for the group
        structure of 3D rotations during the solve.
    \end{itemize}

@@

    Given known vectors in some inertial frame, ${}^N v_i$, and measurements of these
    vectors in some body fixed frame, ${}^B v_i$, our goal is to determine the relative
    rotation from the body to inertial frame ${}^N q^B$, expressed as a quaternion. We
    can define Wahba's loss function as the following: 
    \begin{align} \label{eq:wahba_loss}
        L &= \sum_i w_i \norm{{}^N v_i - {}^N A(q)^B \,\, ^Bv_i }_2^2. = \norm{r_i(q)}_2^2 
    \end{align} 
    where $r_i(q)$ is the residual vector.

    We can solve for ${}^N q^B$ using a nonlinear least squares method minimizing
    Wahba's loss function:
    \begin{mini*}
        {q}{ \norm{r(q)}_2^2 }{}{}
        \addConstraint { q\in SO(3).}
    \end{mini*}

@@

    Alternatively, if we instead minimize with respect to a differential quaternion $\phi$,
    we adapt the algorithm by simply ``correcting'' our Jacobian $J$ using 
    \eqref{eq:quat_jacobian}:
    \begin{equation} 
        \bar{J} = \pdv{r(\mathbf{q} \otimes \varphi(\phi))}{\phi} = \pdv{r(q)}{q} G(q).
    \end{equation}
    We obtain our step---this time in the actual tangent space---as before:
    $\phi = (\bar{J}^T \bar{J})^{-1} \bar{J}^T)$. To obtain our next iterate, we ``add'' the
    step using the correct notion of composition for the group: 
    $\q_{k+1} = \q_k \otimes \varphi(\phi)$. This ``multiplicative'' Newton algorithm is 
    summarized in Algorithm \ref{alg:mgn}.

    \begin{algorithm} 
    	\begin{algorithmic}[1]
    		\caption{Multiplicative Gauss-Newton Method}\label{alg:mgn}
    		\State $k = 0$
    		\While{significant progress}
    		    \State $\bar{J} = \pdv{r(q_k)}{q} G(q_k)$ \Comment{quaternion adjusted Jacobian}
    		    \State $ \phi = -(\bar{J}^T \bar{J})^{-1} \bar{J}^T r(q_k)$ 
    		    \State $\q_{k+1} = \q_k \otimes \varphi(\phi)$ \Comment{apply step multiplicatively}
    		    \State $k = k + 1$
    		\EndWhile
    	\end{algorithmic}
    \end{algorithm}

@@

\section{Trajectory Optimization on $\R^n \times SO(3)$} \label{sec:trajopt}
    Here we outline the modifications to the ALTRO solver \cite{howell2019altro}, to
    solve trajectory optimization problems for rigid bodies, which extends easily to
    arbitrary systems in $\R^n \times SO(3)$. ALTRO is an efficient solver for constrained
    nonlinear optimization problems that uses iterative LQR (iLQR) with an augmented
    Lagrangian framework.

@@

    With this expansion, we calculate the expansion of the ``action-value function'' $Q(x,u)$
    as normal:
    \begin{align}
        Q_{xx} &= \ell_{xx} + A_{k}^T P_{k+1} A_{k} \label{Qxx_exp}\\
        Q_{uu} &= \ell_{uu} + B_{k}^T P_{k+1} B_{k} \label{Quu_exp}\\
        Q_{ux} &= \ell_{ux} + B_{k}^T P_{k+1} A_{k} \label{Qux_exp}\\
        Q_x &= \ell_x  + A_{k}^T p_{k+1} \label{Qx_exp}\\
        Q_u &= \ell_u + B_{k}^T p_{k+1} \label{Qu_exp},
    \end{align}
    from which we can calculate the quadratic expansion of the cost-to-go 
    $P_k \in \R^{12 \times 12}$, $p_k \in \R^{12}$, and optimal linearized feedback gains 
    $K_k \in \R^{m \times 12}$ and $d_k \in \R^m$ by starting at the terminal state and
    resursing backward in time along the trajectory during the ``backward pass'' of the iLQR
    algorithm. During the ``forward pass'', the dynamics are simulated forward in time using
    the feedback gains computed during the backward pass. At each time step, the control is 
    calculated using the linear feedback controller: 
    \begin{equation} \label{eq:mlqr_control}
        u_k = K_k \delta x_k + \bar{u}_k.
    \end{equation}
    where $\bar{u}_k$ is the control value from the previous iteration, and $\delta x$ is
    computed using \eqref{eq:state_error}, with $x_k$ being the current state estimate
    and $\bar{x}_k$ the state from the previous iteration. The rest of the algorithm is
    left unchanged. For more details on the ALTRO algorithm, the reader is encouraged to
    refer to the original paper \cite{howell2019altro}.

@@

\section{Experiments} \label{sec:experiments}
    In this section we present several trajectory optimization problems for systems that
    undergo large changes in attitude: an airplane barrell roll, a quadrotor flip, and a
    satellite with flexible modes that must slew to a new orientation while avoiding a 
    ``keep out'' orientation zone. All results were run on a desktop computer with an 
    AMD Ryzen 2950x processor with 40 GB of RAM. 
    All problems are run using ALTRO, first without any of the modifications presented 
    in the current paper, labeled ``original'', and then using the modifications listed
    in Sec. \ref{sec:trajopt} and the geodesic cost funtion described in Sec. \ref{sec:geodesic},
    labeled ``modified''.
    Timing results are summarized in Table \ref{tab:timing_results}.
    All experiments were solved to a constraint satisfaction tolerance of $10^{-5}$ and 
    discretized with a 4th order Runge-Kutta integrator.
    Code for all experiments is available on 
    \href{https://github.com/RoboticExplorationLab/PlanningWithAttitude}
    {GitHub\footnote{\url{https://github.com/RoboticExplorationLab/PlanningWithAttitude}}}.

@@

        We also compared the ``error quadratic'' cost function described in Section
        \ref{sec:error_quadratic} to the geodesic cost function in Section
        \ref{sec:geodesic}. The geodesic cost was more efficient, both in term of
        iterations and computation time, taking only 36 iterations and 1.8 ms/iteration
        vs 76 iterations and 7.6 ms/iteration for the error quadratic. Not only is the
        geodesic cost function much cheaper to compute, it also converges much faster
        than the geodesic cost function. This could be due to the strong nonlinearities
        introduced when computing the quaternion error, making it more difficult to
        optimize.

@@

	    Figure \ref{fig:quad_flip} shows snapshots of the trajectory as generated using
	    ALTRO. The original version of ALTRO, even after significant tuning efforts,
	    would not converge to the desired solution. It instead got halfway and then
	    ``unwound'' 360 degrees and then continued rotating to the final orientation.
	    This behavior is common and expected when attempting optimization that does not
	    properly account for the group structure of rotations. It is also worth noting
	    that this problem could not be solved using any three-parameter attitude
	    representation, since it passes through the singularities at $90\degree,
	    180\degree$, and $360\degree$ for Euler angles, Rodrigues parameters, and
	    Modified Rodrigues Parameters, respectively.

        \begin{figure}[t]
            \centering
            \includegraphics[width=\columnwidth]{figures/quadflip.png}
            \caption{Snapshots of the quadrotor flip trajectory. The
                gree-colored quadrotors represent the state near t=0 s and the
                red-colored quadrotors represent the state near t=5.0 s
            }
            \label{fig:quad_flip}
        \end{figure}

@@

    With the modifications presented, ALTRO can solve problems few other methods for 
    trajectory optimization can. Many state-of-the-art methods such as direct 
    collocation or sequential convex programming rely on commercial, proprietary, or 
    general-purpose solvers whose internal numerics often are abstracted away from the user. 
    By exploiting both the unique Markovian structure of the trajectory optimization problem
    and the group structure of rotations, ALTRO will likely be able to solve new classes of 
    problems with at or near real-time performance. Future work will focus on continued 
    refinement of the implementation and benchmarking against current state-of-the-art 
    methods for trajectory optimization and other specialized solvers for optimization on
    manifolds.
